# Identifiability of parameters in birth-death models

## First: simulating birth-death trees

So far, we have been treating phylogenetic trees as given
      without asking how they are created (by some
      biologist, presumably). If the traits we are
      interested in do not influence diversification,
      extinction/death, or the probability we observe
      a particular lineage, then the standard Markov models
      on trees seem sufficient. In some cases, traits
      might influence the process creating the phylogenetic
      relationships we observe, and we would like to detect
      these effects.

It is for this reason that we start by considering models for
      tree generation (sometimes called "tree priors" by
      people who use BEAST, which we will get to later in
      the week).

Let's start by simulating some trees using Yule and birth-death models.
We will use the diversitree package:

```{r, message=FALSE, warning=FALSE}
library(diversitree)
```

### Yule process

The **Yule model** is a special case of birth death models. The Yule model is characterized 
by having the death rate set to $\mu=0$. Let's simulate a tree from the Yule process with $\lambda = 1$ for four units of time:

```{r}
tree <- tree.bd(c(lambda=1, mu=0), max.t = 4)
plot(tree, show.tip.label=F)
```

Recall that the probability of $n$ lineages by time $t$, assuming we start with one lineage at time $0$, is given by the solution to the differential equation

$$
\begin{align*}
	\frac{dp_n}{dt} &= \lambda (n-1) p_{n-1} - \lambda n p_n, \quad \text{for } n > 1, \\
	\frac{dp_1}{dt} &= - \lambda p_1.
\end{align*}
$$

This is the Forward Equation (do you remember what the Backward Equation is?).

As a sanity check, let's simulate a large number of Yule trees up to time $t=4$ to see whether the number of lineages
we observe matches the prediction from the Forward Equation. We can easily obtain an approximate solution to the Forward Equation because it is a linear albeit infinite dimensional system of ODEs. This is a little easier to see if we rewrite it as an (infinite-dimensional) matrix equation:

$$
\begin{align*}
	\frac{d}{dt}	
	\begin{pmatrix}
		p_1 \\ p_2 \\ . \\ . \\ . \\ p_n \\ p_{n+1} \\ . 
	\end{pmatrix} 
	 = 
	\begin{pmatrix}
		- \lambda & 0 & 0 & 0 & . & . & . & 0 & 0 & . & .\\
		 \lambda & -2 \lambda & 0 & 0 & . & . & . & 0 & 0 & . & .\\
		 0 & 2 \lambda & -3 \lambda & 0 & . & . & . & 0 & 0 & . & .\\
		 0 & .  & . & . & . & . & . & . & . & . & .\\
		 0 & .  & . & . & . & . & . & . & . & . & .\\
		 0 & .  & . & . & . & . & . & (n-1) \lambda & -n \lambda & . & .\\
		 0 & .  & . & . & . & . & . & . & n \lambda & -(n+1) \lambda   & .\\
		 0 & .  & . & . & . & . & . & . & . & . & .\\
	\end{pmatrix}	
	\begin{pmatrix}
		p_1 \\ p_2 \\ . \\ . \\ . \\ p_n \\ p_{n+1} \\ . 
	\end{pmatrix} 
\end{align*}
$$

If we truncate this system $dP/dt = Q P$ at large but finite $N$, then we can approximate the solution as
$P_N(t) \approx e^{Q_Nt} P_{N}(0)$. The approximation should be reasonable if we choose $N$ sufficiently large. If we simulate only up to time $t=4$, $N$ probably doesn't need to be to huge. If we simulate for longer amounts of time, then we will need make $N$ larger.

```{r}
# How long to simulate?
t <- 4

# Max number of lineages in the approximate solution?
N <- 1000

# Birth rate:
lambda <- 1

# Set up the generator:
Q <- diag(N)
diag(Q) <- - (1:N) * lambda
diag(Q[-1,-N]) <- (1:(N-1))*lambda

# Initial condition: 1 lineage alive at time 0
p0 <- rep(0,N)
p0[1] <- 1

# load in routines to carry out matrix exponentiation for sparse matrices:
require(expm)

# Perform the matrix exponentiation:
pn <- expAtv(Q, p0, t)$eAtv

```

Probability should be conserved. Our truncation does not respect this (the lastlast  column sum of $Q_N$ is negative, meaning there is a net flux of probability out of the truncated system). We can use this to check if our truncation is reasonable -- if we chose $N$ sufficiently large, then the probability flux out of the system should be negligible. That is, the entries of $P_N$ should sum to something very close to one:

```{r}
sum(pn)
```

If we chose $N$ too small in the code at the beginning, then we know that we will be losing probability from the system (which is nonsensical). The large value of $N$ I chose here is ensuring that we are conserving probability, which is good. Experiment with a smaller truncation size $N$ above and re-examine this if it doesn't make sense to you. 

Now let's simulate a large number of Yule trees to compare with the approximate solution from the Master Equation:

```{r}
## Next, we simulate a large number of Yule trees for t units of time:

yulelist <- lapply(1:1000, function(x) tree.bd(c(lambda=1,mu=0), max.t=t))

# Sometimes these fail to make a tree (no births by time t?), so we toss these out:
inds <- sapply(yulelist, is.null)
yulelist <- yulelist[!inds]

n <- sapply(yulelist, Ntip)
```

So now we have simulated lots of Yule trees (~ 1000 of them) and we can compare the distribution of lineages with the prediction from the Master equation:

```{r}
hist(n, freq=F, main='Distribution of lineages by time t=4', breaks=20)
lines(pn, lwd=2)
legend('topright', legend=c('Master equation solution'), 
	lwd=2,lty=1,col='black') 

```

These appear to be in agreement. The approximation will break down if (i) we simulate over a longer period of time, $t$, (ii) if we simulate with a larger birth rate $\lambda$, or (iii) truncate the Master equation at a smaller value of $N$. You should experiment with changing all three of these quantities.

**To summarize**: the solution to the Master equation gives us a way to quantify the probability we observe a Yule tree with $n$ lineages and height $t$. Yule trees that differ only in topology but otherwise have the same height and number of lineages are equally likely to occur in this model.

This is useful for **parameter estimation**. But before we do that, let's look at the birth death model with a positive death rate.

### Birth death process

If the death rate is positive ($\mu > 0$), the process is fundamentally different. This is because the
process can go extinct. 

Let's use the lapply function to generate a
list of simulated outputs (lapply is like a
for loop). Instead of simulating up to time $t$, let's simulate
trees up to a fixed number of taxa -- 50, in our case.

Let's see how frequently the process goes extinct 
before we reach 50 taxa.

```{r}
numtrees <- 50
treelist <- lapply(1:numtrees,
                function(x) tree.bd(c(lambda=2, mu=1), max.taxa = 50)
)

dieoffs <- sapply(treelist, is.null)
numdieoffs <- sum(dieoffs)
```

How many died off before reaching 50 taxa?

```{r}
print(numdieoffs)
```

This is sort of interesting. In the first tree we simulated above, we set $\lambda=1$
and $\mu=0$, for a net birth rate equal to 1. In this second example, we set $\lambda=2$ and
$\mu=1$. The net birth rate still equals 1, but extinction is far more likely now.

**Follow up question**: if we look at a tree produced by $\lambda=1$, $\mu=0$ and 
another tree (successfully) produced with $\lambda=2$, $\mu=1$, are they noticeably different?

```{r}
tree_yule <- tree.bd(c(lambda=1,mu=0),max.taxa=50)
tree_bd <- treelist[!dieoffs][[1]]


par(mfrow=c(1,2))
plot(tree_yule, main=bquote(lambda~'=1, '~mu~'=0'),
	show.tip.label=F)

plot(tree_bd, main=bquote(lambda~'=2, '~mu~'=1'),
	show.tip.label=F)

```

It isn't super obvious whether these trees have different statistical properties. One clue is to examine
the number of lineages through time:

```{r}
require(ape)

par(mfrow=c(1,2))
ltt.plot(tree_yule, xlab='Time', ylab='No. of lineages',
	xlim=c(-5,0),main=bquote(lambda~'=1, '~mu~'=0'))
ltt.plot(tree_bd, xlab='Time', ylab='No. of lineages',
	xlim=c(-5,0),main=bquote(lambda~'=2, '~mu~'=1'))

```

This suggests these trees are different (depending on the random seed used to simulate the trees!). You should try running this several times to see if you notice a pattern.

If you do this, you may notice a tendency for birth-death trees to have shorter heights than their Yule counterparts (conditioning on the same number of taxa, of course). Indeed, conditioning on 50 taxa in the birth death process biases us towards to trees that grow rapidly. The ones that grow more slowly could be at greater risk of extinction. In the Yule process there is no extinction, so we are more likely to observe slowly-growing trees that eventually reach 50 taxa.

## Exercise 
Exercise: Instead of simulating trees up to a fixed number of taxa, simulate trees up to a fixed time. Do you notice any differences between Yule and birth death trees in terms of the number of taxa produced by time $t$?

## Estimating parameters using likelihoods

There is another way to determine whether these trees are statistically indistinguishable or not. We can use
the likelihood function, which quantifies the probability (density) of observing a particular tree given
model parameters. If the trees are statistically indistinguishable, then it should **not** be possible for us to estimate the parameters of the models. If we **can** estimate parameters, then that means the trees contain information about whether the underlying process was Yule or birth-death.

We've been doing Maximum Likelihood a lot up to now. In case you need extra reading material, [here you go](https://en.wikipedia.org/wiki/Maximum_likelihood_estimation).

We've seen a couple of examples of likelihood functions already in this tutorial. The solutions to the Master equations for these processes tell us the probability, $p_n(t)$, of observing $n$ lineages by time $t$. Put another way, this gives us the probability of observing a phylogenetic tree of height $t$ with $n$ taxa. 

The terminology ``likelihood'' means that we consider $p_n(t)$ as a function of the model parameters ($\lambda$ in the Yule process, or $(\lambda, \mu)$ in the birth-death process). Maximum likelihood means that, given our tree of height $t$ with $n$ taxa, we maximize $p_n(t)$ with respect to the parameters of the model. These optimal parmeter values are the Maximum Likelihood Estimate. 


We can create likelihood functions for these trees like so:

```{r}

loglik.yule <- make.bd(tree_yule)
loglik.bd <- make.bd(tree_bd)
```

These don't actually use the Master equation to obtain $p_n(t)$. Instead, they interpret trees as realizations of an [Inhomogeneous Poisson Process](https://en.wikipedia.org/wiki/Poisson_point_process#Inhomogeneous_Poisson_point_process). The underlying theory is the same, however.

Let's try finding the MLE for the Yule tree (supplying the ground truth as the initial guess in the optimization procedure):

```{r}
mle.yule <- find.mle(loglik.yule, c(1,0), method='subplex')
print(mle.yule$message)
print(mle.yule$lnLik)
print(mle.yule$par)
```

Looks pretty close. 

How about for the birth-death tree (again, supplying the ground truth as the initial guess in the optimization procedure):

```{r}
mle.bd <- find.mle(loglik.bd, c(2,1), method='subplex')
print(mle.bd$message)
print(mle.bd$lnLik)
print(mle.bd$par)
```

Seems like it's kind of in the ballpark.

One downside with this is that we are dealing with fairly small trees, and stochasticity is probably introducing lots of noise. If our goal is to understand the types of information we are likely to recover from realizations of these processes, we need to look at a large number of simulations so that statistical regularity kicks in.

## Benchmarking from simulations

Let's simulate more
trees. The make.bd function can be lapply'd to the treelist to give us a
list of likelihood functions corresponding to our list of trees.

```{r}
treelist.yule <- trees(c(lambda=1,mu=0),
		type=c("bd"), n=1500,
		max.taxa=50)

# make a log-likelihood function for each tree:
logliklist <- lapply(treelist.yule, make.bd )

```

What does the distribution of MLEs look like for this set of Yule trees?

```{r}
mles <- lapply(logliklist, function(x){find.mle(x, c(1,0), method='subplex')})

mles_pars <- t(sapply(mles, function(x) x$par))

par(mfrow=c(1,2))
hist(mles_pars[,1],main='Yule tree - estimated birth rate',
	xlab=bquote(hat(lambda)))
hist(mles_pars[,2],main='Yule tree - estimated death rate',
	xlab=bquote(hat(mu)))
```

The estimates $\hat{\lambda}$ seem to be close to the true value of $\lambda=1$, but the
extinction rate, which was $\mu=0$, seems a little harder to pin down.

Remember that the true value of mu we used was mu=0. A lot of the models suggest values of
$\mu$ > 0, which suggests that Yule trees sometimes look birth-death-like. 


Let's try this for birth-death trees next:

```{r}
treelist.bd <- trees(c(lambda=2,mu=1),
		type=c("bd"), n=1500,
		max.taxa=50)

# make a log-likelihood function for each tree:
logliklist <- lapply(treelist.bd, make.bd )

mles <- lapply(logliklist, function(x){find.mle(x, c(2,1), method='subplex')})

mles_pars <- t(sapply(mles, function(x) x$par))

par(mfrow=c(1,2))
hist(mles_pars[,1],main='BD tree - estimated birth rate',
	xlab=bquote(hat(lambda)))
hist(mles_pars[,2],main='BD tree - estimated death rate',
	xlab=bquote(hat(mu)))
```

The birth-death trees with $\lambda=2$, $\mu=1$ clearly have a tendency to convey information about the birth rate $\lambda$ being elevated above the net birth rate $\lambda-\mu=1$. Again, however, the death rate is difficult to pin down.

# Exercises:

## 1.
Use constrain(loglik, mu ~ 0) to
create constrained log-likelihood functions for the Yule trees, and fit them to obtain new estimates for lambda. How does the
variance in $\hat{\lambda}$ change?


## 2.
The Generalized Likelihood Ratio Test is one way to assess whether $\mu$ is significantly different from
zero. The model with more parameters will always fit the data better, but it may not fit the data "that much"
better. The GLR test looks a log(ratio of likelihoods) = difference(log-Likelihoods) to produce a test
statistic (also, rather annoyingly in this case, called $\lambda$).

Use

```{r,eval=F}

constrain(loglik, mu ~ 0)

```

to create the constrained log-likelihood functions, and fit them to obtain new estimates for lambda.

Then formulate a GLR test to assess significant departues from a null hypothesis of $\mu=0$.

(Follow up: the GLR test makes an asymptotic approximation. What is that approximation, and under
what situations would it apply in this case?)

(Hint: find.mle stores log-likelihood values in lnLik).

## 3.
Consider a different hypothesis: $\lambda > \mu$. Look at your estimates $\hat{\lambda}$
and $\hat{\mu}$ to see how often would you conclude that $\lambda < \mu$.

If your simulation produced trees with estimated values of $\hat{\lambda} < \hat{\mu}$, plot these trees and comment
on any features they exhibit that stand out to you. If your simulation did not produce trees with
MLEs having $\hat{\lambda} < \hat{\mu}$, look at a friend's computer to see if theirs did.

## 5.
The find.mle function uses a variety of numerical optimization algorithms to find MLEs. There is also a
function called \textbf{mcmc} that performs Markov Chain Monte Carlo. It accepts a log-likelihood function, and initial parameter guess (x.init),
and a number of iterations to run (nsteps). It also requires a ``tuning parameter'' called w to help the
MCMC routine work more efficiently.

Use the mcmc function with nsteps=1000 and w=c(2,1). Plot loglikelihood ("p")
against parameters to visualize profile log-likelihoods, and also plot lambda against mu. Comment
on what you see, and discuss implications of correlations between lambda and mu. (Note: there are
other mcmc routines besides this one in diversitree. If you have a favorite method, feel free to
use it.)

## 6.
Replace max.taxa with max.t in the tree.bd function to simulate trees
for a fixed amount of time (with variable numbers of taxa), and then repeat Exercises 1-5. (Caution:
injudicious choices of lambda and max.t can make your computer explode. Ctrl + c is useful (on Mac,
anyway) for cancelling calculations in the Terminal.)

## Birth-death-sampling models (and the unidentifiability problem)

In the previous section, we saw that it is in principle possible to estimate paramters of the Yule model and the birth-death model. The statistical behavior of the Maximum Likelihood Estimates might leave something to be desired (especially for the extinction rate $\mu$), but in the limit as the number of taxa tends to infinity, our estimators should be converging in expectation to the [ground truth parameter values](https://en.wikipedia.org/wiki/Consistency_(statistics)).

Models with this property are said to be [identifiable](https://en.wikipedia.org/wiki/Identifiability). The Yule Process and the birth death model are identifiable (assuming that the rates are fixed constants in time). Seemingly innocuous modifications to these models render them unidentifiable. 

The final example we will consider in this tutorial is the birth-death-sampling model. The primary motivation for considering the birth-death-sampling model comes from epidemiology. In that context, pathogens spread between hosts, and infection events correspond to bifurcation events in the pathogen phylogeny. In practice, the vast majority of individuals who are infected are never sampled. Consequently, observed pathogen phylogenies represent a small piece of the true underlying phylogeny. 

In this model, three things can happen:

1. A lineage can split in two (at rate $\lambda$); this corresponds to a new transmission event

2. A lineage can die (at rate $\mu$); this corresponds to either (i) an infected host dying, or (ii) an infected host getting better before transmitting disease

3. A lineage (which resides in an infected host) can be sampled (at rate $\psi$)

We only observe the lineages that get sampled. The phylogeny we observe describes ancestry of the sampled lineages. This is potentially a very small portion of the entire tree.

You might be able to intuit the problem already. We didn't emphasize this in the birth-death model earlier, but in that context we are in fact also observing a subset of the entire phylogeny. The Master equation abstracted away the tree produced by the stochastic process, but in fact it can be helpful to consider dynamic updates to a phylogeny as time proceeds. A birth event corresponds to a lineage bifurcating into two lineages. A death event is somewhat different: when a lienage dies, it gets pruned away from the tree entirely. Thus, a phylogeny produced under a birth-death model is missing all of the lineages that died off before the end of the simulation. 

The heart of the identifiability problem in the birth-death-sampling model is that sampling functions in a very similar way to death. Lineages that are not sampled are pruned away from the tree, just like lineages that die.

If a lineage can either bifurcate at rate $\lambda$, go extinct at rate $\mu$, or be sampled at rate $\psi$, this is a competing rates problem. The probability a particular lineage gets sampled is $\psi/(\lambda+\mu+\psi)$. The probability a lineage dies is $\mu/(\lambda+\mu+\psi)$. You immediately see how to determine the probability a lineage bifurcates before those other two events take place. Some extensions of this model assume lineages can still transmit after sampling, but in this basic model we will assume that sampled lineages are quarantined and treated so that transmission is impossible.

Now, if we simulate birth-death-sampling models and try to estimate parameters, we should expect to run into difficulties.

We will use the castor package to do this.

```{r}
require(castor)
require(ape)

tm <- 4

# Specify model parameters:
lambda <- 3
mu <- 1 
psi <- 1
kappa <- 1

tree.bds = generate_tree_hbds(
  max_time   = tm,          # Maximum duration of the simulation process
  lambda     = lambda,
  mu         = mu,
  psi        = psi,
  kappa      = kappa
)

plot(tree.bds$tree, show.tip.label=F, main='Birth-death-sampling tree')
```

One immediate observation: these trees are generally not ultrametric, as the trees up to now have been. Because we specified a sampling rate, we observe lineages at random times. It is possible to modify the simulation so that sampling takes place at one particular time.

## Exercises

1. Simulate a large number of birth-death-sampling trees and try to replicate our earlier benchmarking analysis to determine if it is possible to simultaneously estimate $\lambda$, $\mu$, and $\psi$. You should probably use the fit_hbds_model_parametric function in castor.
