## Birth death process

Now, we consider the birth-death process. Lineages split at rate $\lambda$, and can die at rate $\mu$.

If the death rate is positive ($\mu > 0$), the process is fundamentally different from the Yule model. This is because the
process can go extinct. 

Let's use the lapply function to generate a
list of simulated outputs (lapply is like a
for loop). Instead of simulating up to time $t$, let's simulate
trees up to a fixed number of taxa -- 50, in our case.

Let's see how frequently the process goes extinct 
before we reach 50 taxa.

```{r}
library(diversitree)

numtrees <- 50
treelist <- lapply(1:numtrees,
                function(x) tree.bd(c(lambda=2, mu=1), max.taxa = 50)
)

dieoffs <- sapply(treelist, is.null)
numdieoffs <- sum(dieoffs)
```

How many died off before reaching 50 taxa?

```{r}
print(numdieoffs)
```

This is sort of interesting. In the first tree we simulated above, we set $\lambda=1$
and $\mu=0$, for a net birth rate equal to 1. In this second example, we set $\lambda=2$ and
$\mu=1$. The net birth rate still equals 1, but extinction is far more likely now.

**Follow up question**: if we look at a tree produced by $\lambda=1$, $\mu=0$ and 
another tree (successfully) produced with $\lambda=2$, $\mu=1$, are they noticeably different?

```{r}
tree_yule <- tree.bd(c(lambda=1,mu=0),max.taxa=50)
tree_bd <- treelist[!dieoffs][[1]]


par(mfrow=c(1,2))
plot(tree_yule, main=bquote(lambda~'=1, '~mu~'=0'),
	show.tip.label=F)

plot(tree_bd, main=bquote(lambda~'=2, '~mu~'=1'),
	show.tip.label=F)

```

It isn't super obvious whether these trees have different statistical properties. One clue is to examine
the number of lineages through time:

```{r}
require(ape)

par(mfrow=c(1,2))
ltt.plot(tree_yule, xlab='Time', ylab='No. of lineages',
	xlim=c(-5,0),main=bquote(lambda~'=1, '~mu~'=0'))
ltt.plot(tree_bd, xlab='Time', ylab='No. of lineages',
	xlim=c(-5,0),main=bquote(lambda~'=2, '~mu~'=1'))

```

This suggests these trees are different (depending on the random seed used to simulate the trees!). You should try running this several times to see if you notice a pattern.

If you do this, you may notice a tendency for birth-death trees to have shorter heights than their Yule counterparts (conditioning on the same number of taxa, of course). Indeed, conditioning on 50 taxa in the birth death process biases us towards to trees that grow rapidly. The ones that grow more slowly could be at greater risk of extinction. In the Yule process there is no extinction, so we are more likely to observe slowly-growing trees that eventually reach 50 taxa.

## Exercises
1. Simulate a large number of Yule and birth-death trees for a fixed number of taxa and compare the distributions of tree heights to determine whether there is in fact a meaningful difference.

2. Instead of simulating trees up to a fixed number of taxa, simulate trees up to a fixed time. Do you notice any differences between Yule and birth death trees in terms of the number of taxa produced by time $t$?

## Likelihood-based inference

There is a principled way to determine whether a tree was produced by one process or the other. We can use
the likelihood function, which quantifies the probability (density) of observing a particular tree given
model parameters. The idea is to use a Generalized Likelihood Ratio test. This works because the Yule model is "nested"
in the birth-death model. Setting $\mu=0$ in the birth-death model produces the Yule model.

The idea behind the GLR test is that, although a more complicated model will always fit data better than a simpler one, it may not fit data "that much" better
if the underlying process is in fact well-described by a simpler model. The GLR test provides way to quantify whether the improvement in fit of a more complicated model is meaningful, or not.
 
We've been doing Maximum Likelihood a lot up to now. In case you need extra reading material, [here you go](https://en.wikipedia.org/wiki/Maximum_likelihood_estimation).

We can create likelihood functions for these trees like so:

```{r}

loglik.yule <- make.bd(tree_yule)
loglik.bd <- make.bd(tree_bd)
```

Let's try finding the MLE for the Yule tree (supplying the ground truth as the initial guess in the optimization procedure):

```{r}
mle.yule <- find.mle(loglik.yule, c(1,0), method='subplex')
print(mle.yule$message)
print(mle.yule$lnLik)
print(mle.yule$par)
```

Looks pretty close. 

How about for the birth-death tree (again, supplying the ground truth as the initial guess in the optimization procedure):

```{r}
mle.bd <- find.mle(loglik.bd, c(2,1), method='subplex')
print(mle.bd$message)
print(mle.bd$lnLik)
print(mle.bd$par)
```

Seems like it's kind of in the ballpark. In the exercises at the end, you are tasked to carry out a GLR test on these trees.


## Benchmarking from simulations

Let's simulate more
trees. The make.bd function can be lapply'd to the treelist to give us a
list of likelihood functions corresponding to our list of trees.

```{r}
treelist.yule <- trees(c(lambda=1,mu=0),
		type=c("bd"), n=1500,
		max.taxa=50)

# make a log-likelihood function for each tree:
logliklist <- lapply(treelist.yule, make.bd )

```

What does the distribution of MLEs look like for this set of Yule trees?

```{r}
mles <- lapply(logliklist, function(x){find.mle(x, c(1,0), method='subplex')})

mles_pars <- t(sapply(mles, function(x) x$par))

par(mfrow=c(1,2))
hist(mles_pars[,1],main='Yule tree - estimated birth rate',
	xlab=bquote(hat(lambda)))
hist(mles_pars[,2],main='Yule tree - estimated death rate',
	xlab=bquote(hat(mu)))
```

The estimates $\hat{\lambda}$ seem to be close to the true value of $\lambda=1$, but the
extinction rate, which was $\mu=0$, seems a little harder to pin down.

Remember that the true value of mu we used was mu=0. A lot of the models suggest values of
$\mu$ > 0, which suggests that Yule trees sometimes look birth-death-like. 


Let's try this for birth-death trees next:

```{r}
treelist.bd <- trees(c(lambda=2,mu=1),
		type=c("bd"), n=1500,
		max.taxa=50)

# make a log-likelihood function for each tree:
logliklist <- lapply(treelist.bd, make.bd )

mles <- lapply(logliklist, function(x){find.mle(x, c(2,1), method='subplex')})

mles_pars <- t(sapply(mles, function(x) x$par))

par(mfrow=c(1,2))
hist(mles_pars[,1],main='BD tree - estimated birth rate',
	xlab=bquote(hat(lambda)))
hist(mles_pars[,2],main='BD tree - estimated death rate',
	xlab=bquote(hat(mu)))
```

The birth-death trees with $\lambda=2$, $\mu=1$ clearly have a tendency to convey information about the birth rate $\lambda$ being elevated above the net birth rate $\lambda-\mu=1$. Again, however, the death rate is difficult to pin down.

# Exercises:

## 1.
Use constrain(loglik, mu ~ 0) to
create constrained log-likelihood functions for the Yule trees, and fit them to obtain new estimates for lambda. How does the
variance in $\hat{\lambda}$ change?


## 2.
The Generalized Likelihood Ratio Test is one way to assess whether $\mu$ is significantly different from
zero. The model with more parameters will always fit the data better, but it may not fit the data "that much"
better. The GLR test looks a log(ratio of likelihoods) = difference(log-Likelihoods) to produce a test
statistic (also, rather annoyingly in this case, called $\lambda$).

Use

```{r,eval=F}

constrain(loglik, mu ~ 0)

```

to create the constrained log-likelihood functions, and fit them to obtain new estimates for lambda.

Then formulate a GLR test to assess significant departues from a null hypothesis of $\mu=0$.

(Follow up: the GLR test makes an asymptotic approximation. What is that approximation, and under
what situations would it apply in this case?)

(Hint: find.mle stores log-likelihood values in lnLik).

## 3.
Consider a different hypothesis: $\lambda > \mu$. Look at your estimates $\hat{\lambda}$
and $\hat{\mu}$ to see how often would you conclude that $\lambda < \mu$.

If your simulation produced trees with estimated values of $\hat{\lambda} < \hat{\mu}$, plot these trees and comment
on any features they exhibit that stand out to you. If your simulation did not produce trees with
MLEs having $\hat{\lambda} < \hat{\mu}$, look at a friend's computer to see if theirs did.

## 5.
The find.mle function uses a variety of numerical optimization algorithms to find MLEs. There is also a
function called \textbf{mcmc} that performs Markov Chain Monte Carlo. It accepts a log-likelihood function, and initial parameter guess (x.init),
and a number of iterations to run (nsteps). It also requires a ``tuning parameter'' called w to help the
MCMC routine work more efficiently.

Use the mcmc function with nsteps=1000 and w=c(2,1). Plot loglikelihood ("p")
against parameters to visualize profile log-likelihoods, and also plot lambda against mu. Comment
on what you see, and discuss implications of correlations between lambda and mu. (Note: there are
other mcmc routines besides this one in diversitree. If you have a favorite method, feel free to
use it.)

## 6.
Replace max.taxa with max.t in the tree.bd function to simulate trees
for a fixed amount of time (with variable numbers of taxa), and then repeat Exercises 1-5. (Caution:
injudicious choices of lambda and max.t can make your computer explode. Ctrl + c is useful (on Mac,
anyway) for cancelling calculations in the Terminal.)

## Birth-death-sampling models (and the unidentifiability problem)

In the previous section, we saw that it is in principle possible to estimate paramters of the Yule model and the birth-death model. The statistical behavior of the Maximum Likelihood Estimates might leave something to be desired (especially for the extinction rate $\mu$), but in the limit as the number of taxa tends to infinity, our estimators should be converging in expectation to the [ground truth parameter values](https://en.wikipedia.org/wiki/Consistency_(statistics)).

Models with this property are said to be [identifiable](https://en.wikipedia.org/wiki/Identifiability). The Yule Process and the birth death model are identifiable (assuming that the rates are fixed constants in time). Seemingly innocuous modifications to these models render them unidentifiable. 

The final examples we will consider in this tutorial is the birth-death-sampling model and the birth-death-sampling-through-time model (say that three times in a row). 

The birth-death-sampling model is just a birth-death model, with the extra feature that we only observe a fraction of the tips at the end. In these trees, tips are observed at the same point in time, and so the trees are ultrametric.

The birth-death-sampling-through-time model accommodates a rate of sampling through time. The trees produced by this process are heterochronous, i.e. not ultrametric.

The primary motivation for considering the birth-death-sampling-through-time model comes from epidemiology. In that context, pathogens spread between hosts, and infection events correspond to bifurcation events in the pathogen phylogeny. Infected people are also observed over a period of time (perhaps as they show up at the hospital).

In the sampling-through-time model, three things can happen:

1. A lineage can split in two (at rate $\lambda$); this corresponds to a new transmission event

2. A lineage can die (at rate $\mu$); this corresponds to either (i) an infected host dying, or (ii) an infected host getting better before transmitting disease

3. A lineage (which resides in an infected host) can be sampled (at rate $\psi$)

These three rates appear in the Backward Kolmogorov Equation to calculate the tree likelihood. In the ultrametric case (sampling occurs at one point in time), the Backward Kolmogorov Equation looks like the birth-death model, but the initial conditions at the tips incorporate sampling probabilities.

The most important aspect of these processes is that we only observe the lineages that get sampled. The phylogeny describes ancestry of these sampled lineages. This is potentially a very small portion of the entire tree.

You might be able to intuit the problem already. We didn't emphasize this in the birth-death model earlier, but in that context we are in fact also observing a subset of the entire phylogeny. If we simulate a birth-death model up to time $t$, we never observe lineages that die out before time $t$. The Master equation abstracted away the tree produced by the stochastic process, but in fact it can be helpful to consider dynamic updates to a phylogeny as time proceeds. A birth event corresponds to a lineage bifurcating into two lineages. A death event is somewhat different: when a lineage dies, it gets pruned away from the tree entirely. Thus, a phylogeny produced under a birth-death model is missing all of the lineages that died off before the end of the simulation. 

The heart of the identifiability problem in the birth-death-sampling model is that sampling functions in a very similar way to death. Lineages that are not sampled are pruned away from the tree, just like lineages that die. It turns out that this is a huge problem when sampling occurs at one point in time (ultrametric trees), but is less of a problem when sampling occurs through time (heterochronous trees).

We will use the castor package to see if this for ourselves.

```{r}
require(castor)
require(ape)

set.seed(21)

lambda_true = 0.5
mu_true     = 0.3

## Regime 1: sampling through time
sim_through_time = generate_tree_hbds(
  max_time   = 20,
  lambda     = lambda_true,
  mu         = mu_true,
  psi        = 0.15,   # continuous sampling rate
  kappa      = 0,      # sampled lineages removed from pool (e.g. epi-style)
  include_extant  = FALSE,
  include_extinct = FALSE
)

## Regime 2: sampling only at the end (standard reconstructed tree)
sim_end_only = generate_tree_hbds(
  max_time   = 20,
  lambda     = lambda_true,
  mu         = mu_true,
  psi        = 0,
  CSA_times  = 20,
  CSA_probs  = 0.5,    # rho-style sampling fraction
  CSA_kappas = 1,
  include_extant  = FALSE,
  include_extinct = FALSE
)


par(mfrow=c(1,2))

plot(sim_through_time$tree,show.tip.label=F, 
	main='Sampling thru time')
plot(sim_end_only$tree,show.tip.label=F,
	main='Sampling at the end')
```


## Exercises

1. Simulate a large number of trees under each of the two varieties of models (sampling at the end vs. sampling through time). Can you tell whether the model with sampling at the end is identifiable or not? Again, it may be helpful to break into groups to work on this problem. 
