# The Yule Model 

## First: simulating Yule trees

So far, we have been treating phylogenetic trees as given
      without asking how they are created (by some
      biologist, presumably). If the traits we are
      interested in do not influence diversification,
      extinction/death, or the probability we observe
      a particular lineage, then the standard Markov models
      on trees seem sufficient. In some cases, traits
      might influence the process creating the phylogenetic
      relationships we observe, and we would like to detect
      these effects.

It is for this reason that we start by considering models for
      tree generation (sometimes called "tree priors" by
      people who use BEAST, which we will get to later in
      the week).

Let's start by simulating some trees using Yule model.
We will use the diversitree package:

```{r, message=FALSE, warning=FALSE}
library(diversitree)
```

### Yule process

The **Yule model** is a special case of birth death models. The Yule model is characterized 
by having the death rate set to $\mu=0$. Let's simulate a tree from the Yule process with $\lambda = 1$ for four units of time:

```{r}
tree <- tree.bd(c(lambda=1, mu=0), max.t = 4)
plot(tree, show.tip.label=F)
```

Recall that the probability of $n$ lineages by time $t$, assuming we start with one lineage at time $0$, is given by the solution to the differential equation

$$
\begin{align*}
	\frac{dp_n}{dt} &= \lambda (n-1) p_{n-1} - \lambda n p_n, \quad \text{for } n > 1, \\
	\frac{dp_1}{dt} &= - \lambda p_1.
\end{align*}
$$

This is the Forward Equation (do you remember what the Backward Equation is?).

As a sanity check, let's simulate a large number of Yule trees up to time $t=4$ to see whether the number of lineages
we observe matches the prediction from the Forward Equation. We can easily obtain an approximate solution to the Forward Equation because it is a linear albeit infinite dimensional system of ODEs. This is a little easier to see if we rewrite it as an (infinite-dimensional) matrix equation:

$$
\begin{align*}
	\frac{d}{dt}	
	\begin{pmatrix}
		p_1 \\ p_2 \\ . \\ . \\ . \\ p_n \\ p_{n+1} \\ . 
	\end{pmatrix} 
	 = 
	\begin{pmatrix}
		- \lambda & 0 & 0 & 0 & . & . & . & 0 & 0 & . & .\\
		 \lambda & -2 \lambda & 0 & 0 & . & . & . & 0 & 0 & . & .\\
		 0 & 2 \lambda & -3 \lambda & 0 & . & . & . & 0 & 0 & . & .\\
		 0 & .  & . & . & . & . & . & . & . & . & .\\
		 0 & .  & . & . & . & . & . & . & . & . & .\\
		 0 & .  & . & . & . & . & . & (n-1) \lambda & -n \lambda & . & .\\
		 0 & .  & . & . & . & . & . & . & n \lambda & -(n+1) \lambda   & .\\
		 0 & .  & . & . & . & . & . & . & . & . & .\\
	\end{pmatrix}	
	\begin{pmatrix}
		p_1 \\ p_2 \\ . \\ . \\ . \\ p_n \\ p_{n+1} \\ . 
	\end{pmatrix} 
\end{align*}
$$

If we truncate this system $dP/dt = Q P$ at large but finite $N$, then we can approximate the solution as
$P_N(t) \approx e^{Q_Nt} P_{N}(0)$. The approximation should be reasonable if we choose $N$ sufficiently large. If we simulate only up to time $t=4$, $N$ probably doesn't need to be to huge. If we simulate for longer amounts of time, then we will need make $N$ larger.

For comparison, we can also use the analytical solution $P_n(t) = e^{-\lambda t} (1-e^{-\lambda t})^{n-1}$.


```{r}
# How long to simulate?
t <- 4

# Max number of lineages in the approximate solution?
N <- 1000

# Birth rate:
lambda <- 1

# Set up the generator:
Q <- diag(N)
diag(Q) <- - (1:N) * lambda
diag(Q[-1,-N]) <- (1:(N-1))*lambda

# Initial condition: 1 lineage alive at time 0
p0 <- rep(0,N)
p0[1] <- 1

# load in routines to carry out matrix exponentiation for sparse matrices:
require(expm)

# Perform the matrix exponentiation:
pn <- expAtv(Q, p0, t)$eAtv

```

Probability should be conserved. Our truncation does not respect this (the lastlast  column sum of $Q_N$ is negative, meaning there is a net flux of probability out of the truncated system). We can use this to check if our truncation is reasonable -- if we chose $N$ sufficiently large, then the probability flux out of the system should be negligible. That is, the entries of $P_N$ should sum to something very close to one:

```{r}
sum(pn)
```

If we chose $N$ too small in the code at the beginning, then we know that we will be losing probability from the system (which is nonsensical). The large value of $N$ I chose here is ensuring that we are conserving probability, which is good. Experiment with a smaller truncation size $N$ above and re-examine this if it doesn't make sense to you. 

Now let's simulate a large number of Yule trees to compare with the approximate solution from the Master Equation:

```{r}
## Next, we simulate a large number of Yule trees for t units of time:

yulelist <- lapply(1:1000, function(x) tree.bd(c(lambda=1,mu=0), max.t=t))

# Sometimes these fail to make a tree (no births by time t?), so we toss these out:
inds <- sapply(yulelist, is.null)
yulelist <- yulelist[!inds]

n <- sapply(yulelist, Ntip)

# Let's also create a fucntion for our analytical solution for comparison:
getpn <- function(n) exp(-lambda*t)*(1-exp(-lambda*t))^(n-1)

pn.a <- sapply(0:400, getpn)

```

So now we have simulated lots of Yule trees (~ 1000 of them) and we can compare the distribution of lineages with the prediction from the Master equation:

```{r}
hist(n, freq=F, main='Distribution of lineages by time t=4', breaks=20)
lines(pn, lwd=2)
lines(pn.a, lwd=3, lty=3, col='red')
legend('topright', legend=c('Numerical approximation', 'Analytical solution'), 
	lwd=3,lty=c(1,3),col=c('black','red'), title='Master Equation solution') 

```

These appear to be in agreement. The approximation will break down if (i) we simulate over a longer period of time, $t$, (ii) if we simulate with a larger birth rate $\lambda$, or (iii) truncate the Master equation at a smaller value of $N$. You should experiment with changing all three of these quantities.

**To summarize**: the solution to the Master equation gives us a way to quantify the probability we observe a Yule tree with $n$ lineages and height $t$. Yule trees that differ only in topology but otherwise have the same height and number of lineages are equally likely to occur in this model.

This is useful for **parameter estimation**. But before we do that, let's look at the birth death model with a positive death rate in the next lecture.

# Exercises
1. We can define a likelihood function and fit to a tree as follows:

```{r}
lik.yule <- make.yule(tree)
fit.yule <- find.mle(lik.yule, x.init = 0.1)
coef(fit.yule)      # lambda estimate
logLik(fit.yule)
```

Carry out a simulation study to determine how well we are able to estimate $\lambda$. Specifically, it would be good to find out whether our ability to estimate $\lambda$ is improved by having more tips in our tree. Simulate trees for fixed numbers of taxa. Do you notice any correlations between tree height and $\hat{\lambda}$? It may be helpful to break up into groups.
