# Simulating stochasting character mappings

```{r, message=FALSE, warning=FALSE}
# rm(list=ls())

library(ape)
library(expm) # matrix exponential
library(ggtree)
library(ggplot2)
library(dplyr)
library(cowplot)
library(scatterpie)
library(phytools)
```

Carrying on from our last example of Seasonal Coronavirus NL63, we seek to address the following questions:

- 1. How often did the S amino acid mutate into P, and vice versa?
- 2. Where precisely do we think mutations occurred on the phylogeny?

With ancestral character estimation (ace), we approached answers to these questions. We considered two types of Markov models (equal rates, ER, and all rates different, ARD). The MLEs of the model parameters carry pretty different interpretations, but the ancestral estimates of past ancestors are largely robust. The main difference we saw was concerning the inferred root state of the phylogeny in each model.

Stochastic character mapping, or character mapping, uses our fitted Markov chain models to annotate phylogenies with mutations along edges. So far, we have just been annotating internal nodes of the phylogenies with their most likely states, given the models and the data (tip data + phylogeny).

To perform character mapping, we will make use of a suite of functions from the phytools package in R. 

Let's start by loading in our Coronavirus NL63 phylogeny:

```{r}
tree <- read.tree('coronavirus_nl63/nextstrain_seasonal-cov_nl63_timetree.nwk')

# Root the tree (not the most technically appropriate way to do this, but good enough for our illustrative purposes):
tree <- root(tree, node=Ntip(tree)+1 )

# Make the tree fully bifurcating:
tree <- multi2di(tree)

tree$edge.length[tree$edge.length < 0] <- .01
tree$edge.length[tree$edge.length == 0] <- .01


# Some of these tips have an amino acid change in the gene for the Spike protein. From Nextstrain, we can 
# see that it is these ones:

Stips <- tree$tip.label
Ntips <- setdiff(extract.clade(tree, 402)$tip.label, extract.clade(tree, 473)$tip.label)
Stips <- setdiff(tree$tip.label, Ntips)

x <- rep(0, Ntip(tree))
names(x) <- tree$tip.label
x[Stips] <- "S"
x[Ntips] <- "P"

p <- ggtree(tree) + ggtitle('Seasonal Coronavirus NL63') +
	theme(plot.title=element_text(hjust=0.5, size=18, face="bold"))

p <- p %<+% data.frame(label=tree$tip.label, aminoacid = as.factor(x))

p + geom_tippoint(aes(col=aminoacid))
```

Stochastic character mapping involves 3 steps:

- 1. Fit a Markov model to the phylogeny (we did this using ace already)
- 2. Sample ancestral state at internal nodes using the scaled likelihoods (calculated in step 1)
- 3. Use the estimates from the fitted model to simulate changes of state along edges of the phylogeny that are compatible with the node states at the two ends of each edge (similar to how we simulated event times in the previous tutorial) 

That's it. Conceptually, it's super simple. Computationally, it can be a little bit intense to simulate changes of state along edges that are compatible with node states at either end of the edge.

The phytools package has a few functions that we will use. 

**One of these is fitMk**. This function is similar to ace, except that  user may supply a "root prior" (e.g. 50% probability that the root is in state S or P, for our example). When we supply a root prior, we are optimizing a different objective function than ace. By default, fitMk uses scaled likelihoods at the root, identical to ace. The second distinction between fitMk and ace is that they use slightly different optimizers to find Maximum Likelihood Estimates (MLEs). Sometimes it can be convenient to try using both functions if one of them encounters numerical problems during optimization. It can also be a good idea to obtain estimates from both methods as a sanity check. 

The other useful function is **simmap**, which simulates a stochastic character mapping on a tree given a Markov model generator matrix. This is the function that actually carries out steps 2-3 above.

A third function **make.simmap**, does both steps (fitMk and simmap) in one go. Obviously it is not necessary to invoke make.simmap after running ace or fitMk.

## Comparing fitMk and ace

## Simulate character mappings with simmap
