# Simulating event times

An essential aspect of stochastic character mapping is simulating events
in Markov chains. We have seen that we can calculate transition probabilities
along branches of a phylogeny, and that these can be used to determine the
the probability of particular configurations of states at the internal nodes
and tips of a tree. Armed with a Markov model and the transition probabilities
that can be calculated from its generator, we can simulate configurations of
states at internal nodes.

## Exercises
## 1.

The strategy for stochastic character mapping relies on simulating changes of state 
for Markov chains. So far, we have seen how to calculate transition probabilities given
a fixed amount of time. 
 
To illustrate this idea, let's begin with a simple 2-state Markov chain.

```{r}
Q <- matrix(c(-1,100,1,-100), nrow=2, byrow=T)
```

This generator says that if we are in state 1, we move to state 2 at a rate equal to 1. If we are in state 2, we move back to 
state 1 at a rate equal to 100. So, we expect to be spending more time in state 1 than in state 2. In fact, the relative amount
of time we expect to reside in either state is given by the stationary distribution

$\pi = (\frac{100}{101},\frac{1}{101})^T$.

The output of this simulation is going to be a list/vector/whatever of event times. We start at time zero. I'm going to decide at the outset that I want to
simulate 50 changes of state, so we will have 50 event times.

```{r}
n <- 50

events <- rep(0, n)
states <- rep(1,n)

```


We have to decide whether to start in state 1 or state 2. I have declared that
we start in state 1 above.

Because the rates of our stochastic rate matrix Q are constant in time,
the time until each event is described by an exponential distribution.
The procedure we are about to implement is known as **Gillespie's Algorithm**.
The process we are about to simulate is also a type of **Inhomogeneous Poisson Process**.

The algorithm proceeds as follows. 
If we are currently in state 1, we know that we switch to state 2 at rate $q_{21}$. In the stochastic 
setting, the amount of time until the state change is an Exponentially-distributed random variable with 
rate $q_{21}$, so that $T_{21} \sim q_{21} e^{-q_{21}t}$. If we
are currently in state 2, the time until a transition to state 1 follows
$T_{12} \sim q_{12} e^{-q_{12}t}$.

## 2.
Which convention are we using for $q_{ij}$? Is this the transition rate from state $i$ to state $j$, or the
rate into state $i$ from state $j$?

Let's proceed to simulate this Markov chain in continuous time:

```{r}
for(i in 2:n){
	state=states[i-1]
	events[i] <- events[i-1] + rexp(1, rate= -Q[state, state])

	states[i] <- if(state==1){2}else{1}
}

dat <- cbind(events, states)
plot(states~events,dat,type='s',xlab='Time',ylab='State')
```

There are a couple of takeaways from this plot. First, we see that the system is switching between
states 1 and 2 at random times. Second, the system appears to be spending more time in state 1 than in
state 2. Let's verify that this matches the stationary distribution. 

```{r}
# What was the total simulation time?
Ttot <- tail(dat,n=1)[1]

# Calculate the amounts of time spent in each state:

diffdat <- cbind(diff(dat[,'events']),dat[-dim(dat)[1],'states']) 

# How much total time in state 1?
T1 <-  sum(diffdat[diffdat[,2]==1,1])

# How much total time in state 2?
T2 <-  sum(diffdat[diffdat[,2]==2,1])

print(c(T1/Ttot, T2/Ttot))

```

The fraction of time spent in state 2 should be 1/101 $\approx 0.0099$. We are simulating over a fairly short
chunk of time, so we expect some randomness in any given simulation. 
If we simulate for a longer period of time, these time averages should approach the stationary distribution. 
Try changing n above to a large number (like 10000) to confirm this.

## 3.

Modify the event time simulation above to run for a pre-specified amount of time (or branch length)
and an undetermined number of events. Simulate realizations of the stochastic process that
accomodate predefined initial and final states, or "boundary data". (Hint: you may include a step that invovles rejecting
a trajectory if it does not match with your boundary data.)

## 4.
In simulating a stochastic character mapping, does it matter which direction we simulate from? Mathematically,
there is no reason why we couldn't use tip states as initial conditions, but that doesn't really seem
consistent with evolution occurring in forward time. Are there situations where the two approaches are
equivalent? Why or why not?

