# Simulating nucleotide substitution models
## A refresher on continuous time discrete state Markov chains

First, let's load in some packages we will need -- ape, for working with phylogenetic trees, 
and expm, which calculates matrix exponentials (useful for solving differential equations).

```{r}
rm(list=ls())
library(ape)
library(expm) # matrix exponential
```

# 1. Basic Markov chains

To reconstruct phylogenetic trees from DNA sequence data, we commonly use continuous-time 
discrete-state Markov chains to model changes in nucleotides over time. In the simplest case, we consider
the evolution of a particular site in a genome independent of all the rest. This page is focused on the mathematics
of Markov chains. By understanding how these stochastic processes work, we will be in a better position to understand
the underlying assumptions of many methods in statistical phylogenetics.

Let's begin by considering a simple 2-state Markov chain
defined in continuous time. A site in a DNA sequence is usually modeled with four sites (A, C, T, G). We will 
start with a 2-state Markov chain to learn the basics in the simplest possible case. Extending to four states
is not conceptually different, but becomes mathematically much more cumbersome. 

Let's begin! We consider a vector, $P$, with two entries that sum to one. The components of $P$
are the probability the state resides in one state or the
other at a particular time. The process is simple: if the system is in state 1, it can transition to state 2 
at rate $q_{12}$. If the system is in state 2, it can transition to state 1 at rate $q_{21}$. Easy! But how do we
translate this into a formal mathematical statement? We use differential equations!

The probability the chain is one state or the other changes over time according to the ordinary differential equations
(sometimes called the Kolmogorov equations) 

\begin{align*}
 dP_1/dt &= - q_{12} P_1 + q_{21} P_2,\\
 dP_2/dt &= q_{12}P_1 - q_{21} P_2.
\end{align*}

If we treat P as a column vector,
this can be written in vector-matrix form
\begin{align*}
	dP/dt = Q P,
\end{align*}
where Q is a stochastic rate matrix with columns summing to zero.

In some cases, the row vector $P^T$ is used instead. In this case,
the equation takes the form $dP^T/dt = P^T Q^T$. Pay attention to the
help pages in the coming days to ensure that $Q_{ij}$ represents the
state change that you think it does...

We can solve these differential equations by hand, using the fact
that $P_1 + P_2 = 1$:

\begin{align*}
dP_1/dt &= -q_{12}(P_1) + q_{21}  (1-P_1), \\
	&= - (q_{12} + q_{21})  P_1 + q_{21}.
\end{align*}

This rearranges to:
\begin{align*}
dP_1/dt + (q_{12}+q_{21})  P_1 = q_{21}.
\end{align*}

A standard approach to solving this type of equation is to use an
integrating factor, in this case of the form $e^{(q_{12}+q_{21})t}$, which gets multiplied
on both sides of the equation:
\begin{align*}
e^{(q_{12}+q_{21})t} (dP_1/dt + (q_{12}+q_{21})  P_1 ) = e^{(q_{12}+q_{21})t} q_{21}.
\end{align*}

From Calculus, we can readily recognize that the left-hand-side of this equation looks like the result of a "product-rule" 
operation. We can rewrite as a derivative:

\begin{align*}
d/dt (P_1  e^{(q_{12}+q_{21})t}) = q_{21}  e^{(q_{12}+q_{21})t}.
\end{align*}

This is why we multiplied both sides by the integrating factor -- we're just using differentation rules from Calculus to 
solve the problem. We now have an equation that we can integrate on both sides of the = sign.

Integrating this up once gives 
\begin{align*}
P_1  e^{(q_{12}+q_{21})t} = \frac{q_{21}}{q_{12}+q_{21}}  e^{(q_{12}+q_{21})t} + C,
\end{align*}
where $C$ is a constant of integration we need to determine eventually. We can now isolate P_1:

\begin{align*}
P_1 =  \frac{q_{21}}{q_{12}+q_{21}}  + C e^{-(q_{12}+q_{21})t}.
\end{align*}

We still have to determine $C$. We use initial conditions to do so. Set $P_{1,0} = P_1(0)$:

\begin{align*}
P_{1,0} &= \frac{q_{21}}{q_{12}+q_{21}}  + C e^{-(q_{12}+q_{21})0} ,\\
&= \frac{q_{21}}{q_{12}+q_{21}}  + C,
\end{align*}
so that $C = P_{1,0} - \frac{q_{21}}{q_{12}+q_{21}}$. We get to decide on $P_{1,0}$, which is the probability that the system starts in
state 1 at time 0.


So now we know that
\begin{align*}
P1 &=  \frac{q_{21}}{q_{12}+q_{21}}  + (P_{1,0} - \frac{q_{21}}{q_{12}+q_{21}})  e^{-(q_{12}+q_{21})t},\\
 &= P_{1,0}  e^{-(q_{12}+q_{21})t} + \frac{q_{21}}{q_{12}+q_{21}}  (1-e^{-(q_{12}+q_{21})t})
\end{align*}

Notice that this has a term corresponding to the initial condition that
is dominant for $t \approx 0$, and in the long run, as $t \to \infty$,  $P_1 \to \frac{q_{21}}{q_{12}+q_{21}}$.

We can also find $P_2$, using the fact that $P_2 = 1-P1$:
\begin{align*}
P_2 &= \frac{q_{12}+q_{21}}{q_{12}+q_{21}} - \frac{q_{21}}{q_{12}+q_{21}} + \frac{q_{21}}{q_{12}+q_{21}}e^{-(q_{12}+q_{21})t} - P_{1,0} e^{-(q_{12}+q_{21})t},\\
 &= \frac{q_{12}}{q_{12}+q_{21}} + (\frac{q_{21}}{q_{12}+q_{21}} - P_{1,0})e^{-(q_{12}+q_{21})t},\\
 &= \frac{q_{12}}{q_{12}+q_{21}} + (\frac{q_{21}}{q_{12}+q_{21}} - 1 + P_{2,0})e^{-(q_{12}+q_{21})t},\\
 &= \frac{q_{12}}{q_{12}+q_{21}} + (\frac{-q_{12}}{q_{12}+q_{21}} + P_{2,0})e^{-(q_{12}+q_{21})t},\\
 &= \frac{q_{12}}{q_{12}+q_{21}}(1-e^{-(q_{12}+q_{21})t}) + P_{2,0}e^{-(q_{12}+q_{21})t}.
\end{align*}
So now, we have our solutions for the Markov chain probabilities over time, which we can define in a handy R function:

```{r}
P <- function(t, P0, q12,q21){
	P1 <- function(t){
		P0[1]*exp(-(q12+q21)*t) + q21/(q12+q21) * (1-exp(-(q12+q21)*t))
	}
	P2 <- function(t){
		P0[2]*exp(-(q12+q21)*t) + q12/(q12+q21) * (1-exp(-(q12+q21)*t))
	}
	P <- c(P1(t), P2(t))
	return(P)
}
```

# Exercises:
## 1. 
Use the function above to plot the probability that the system begins in state 1
	and transitions to state 2 as a function of time. The initial condition in this
	case can be specified by setting ```P0 <- c(1,0)```.
	You will have to decide
	on values of $q_{12}$ and $q_{21}$. What do you notice if you make $q_{12}$ 
	large? Does a state transition from state 1 to 2 happen faster, or slower? 
	Does the probability of a state transition from state 1 to state 2 depend on $q_{21}$ at all?	


## 2.
Write down a 4-state Markov model corresponding to A, C, T, and G
	in DNA. Evaluate the transition probabilities as above (hint:
	the matrix exponential can be calculated in R using
	the expm function from the package of the same name. Feel free
	to assume whatever constraints on parameters you like (or none
	at all).

## 3.
The Jukes-Cantor model is the simplest nucleotide substitution model.
	By assumption, the equilibrium probabilities of each nucleotide
	are equal, and the transition rates between states are likewise
	all equal. What is the generator for this process? Write a function
	to describe transition probabilities as a function of time.
	(Hint: how many parameters will your generator have
	if all transitions between states occur at equal rates?)

## 4.
The K80 (Kimura, 1980) model elaborates on the Jukes-Cantor model by
	allowing substitutions within pyrimidines (C,T) and within purines
	(A, G) to occur at a different rate than "transversions" between
	 pyrimidines and purines. In biology, substitutions between two
	purines or between two pyrimidines are called "transitions" (but
	in Markov chains, mathematicians refer to all state changes as
	transitions -- confusing). Assuming that transitions (in the
	biological sense of the word) occur at the same rate regardless of
	whether the nucleotides are purines or pyrimidines, and assuming
	that transversions in either direction occur at the same rate,
	write the generator for the Markov process and write a function that
	calculates transition probabilities as a function of time.


