# Intro to trees in R

This page is basically following the vignette Ancestral.html for the phangorn package in R.

The original vignette is written by Klaus Schliep (klaus.schliep@gmail.com).

I'm annotating it a bit more for our purposes.

First, let's install some of the packages we will need:

```{r, warning=FALSE, message=FALSE}
library(phangorn)
library(phytools)
library(ape)
```


Next, let's load in some data. We will read in an example dataset from the phangorn package in R,
which contains genetic sequence information about a number of different species.

```{r}
#Let's load in some primate data:
fdir <- system.file("extdata/trees", package = "phangorn")
primates <- read.phyDat(file.path(fdir, "primates.dna"),
                        format = "interleaved")
```

Let's examine this data file - what does it contain?

```{r}
names(primates)
```

This shows us there are 14 sequences corresponding to 14 different species. Let's look at Mouse:

```{r}
primates$Mouse
```

It's a sequence of the numbers 1--4.
Typing primates into the console indicates that the states are a,c,g,t:

```{r}
primates
```

So the Mouse sequence is really

```{r}
c('a','c','t','g')[primates$Mouse]
```

The letters "a", "c", "t", and "g" correspond to the different nucleotides in deoxyribonucleic acid (DNA). If you've never
heard of DNA, say hello to the bugs (the ones living there under the rock with you).

Let's turn this species data into an R data frame to visualize the sequence alignment. Columns correspond to species, and rows
correspond to a position in the DNA sequence:

```{r}
primdat <- as.data.frame(primates)
```

Let's look at the the beginning of the multiple sequence alignment:

```{r}
head(primdat)
```

The sequences appear to be mostly similar. If we had included bugs or plants in the data, there might
be more obvious differences. The fact that the sequences are rather similar among these species indicates
that they share ancestry somewhat recently (at least in the grand scheme of life on Earth as we know it). 

Let's look at each pair of species in the data
and find the number of differences in their sequences.

For example, if we wanted to see how many sites differ between Mouse and Human, we could
do this:

```{r}
sum(primdat$Mouse != primdat$Human)
```

128 sites differ between these species; there are

```{r}
dim(primdat)[1]
```

total sites in each sequence.

We will load these into a matrix, where the i,j entry
tells us how many sites differ between species i and species j.

This matrix will obviously be symmetric, and the diagonal will be zero. S
So, we can can just fill in the upper diagonal of the matrix to get the information we need, 
and then use symmetry to fill in the lower half:

```{r}
seq_diff_matrix <- matrix(nrow=14,ncol=14)
colnames(seq_diff_matrix) <- names(primates)
rownames(seq_diff_matrix) <- names(primates)

diag(seq_diff_matrix) <- 0

#I will swallow my pride and write a nested for loop (real R pros use the apply family of functions):
for(i in 1:14){
	for(j in i:14){
		seq_diff_matrix[i,j] <- sum( primdat[,i] != primdat[,j] )
	}
}

seq_diff_matrix[lower.tri(seq_diff_matrix)] <- t(seq_diff_matrix)[lower.tri(t(seq_diff_matrix))]

#check to make sure I did this right -- is the matrix equal to its transpose? (i.e. is it symmetric?)
seq_diff_matrix == t(seq_diff_matrix)

#Yes it is -- now plot!
heatmap(seq_diff_matrix, symm=T)

```

So already, it is abundantly clear that Humans, Chimps, and Gorillas
are are pretty similar in terms of their DNA sequences. There is also a group
in the middle (BarbMacq, Crab-E.Mac, Jpn Macaq, and Rhesus Mac) that look pretty similar to each other.

The heatmap function even makes a nice little dendogram tree on the margins!

We have made a tree, but there are some potential drawbacks with this. The tree should reflect the evolutionary history of these organisms.

That means the internal nodes of the tree should have their own sequences because they correspond to ancestral organisms 
with their own DNA sequences. In constructing
a phylogenetic tree, we should be taking account of the fact that closely related things have shared ancestry, and the ancestor will look
similar to its recent descendants. 

In this tree we have produced so far, we have only looked at raw differences between pairs of sequences. We have
not explicitly made use of the idea of "common descent" from ancestors. The idea of similarity due to common descent means that 
the location of nucleotide differences in the sequences will matter. Internal nodes of our phylogenetic tree should correspond to ancestors
of present-day organisms. These ancestors will have nucleotide sequences of their own, and these should be similar to the sequences in the
descendants. We have not accounted for this.

This is why the method of Parsimony was originally developed. Instead of looking at pairwise differences
between sequences, we try to determine what the evolutionary history would have been, under the
assumption that evolution is "parsimonious" -- the number of nucleotide changes from the root of the tree
to the leaves should be minimized. 

Let's calculate a tree using Parsimony:


```{r}
tree <- pratchet(primates, trace=0) |> acctran(primates) |> makeNodeLabel()
 parsimony(tree, primates)

anc.pars <- anc_pars(tree, primates)

#Let's look at the evolution of site 17 in the sequences:
plotAnc(anc.pars, 17)
title("Maximum Parsimony Reconstruction")
```

The Parsimony reconstruction method provides us with a way to estimate what ancestral sequences might
have looked like. Already, we can see that species that are closely related tend to have the same sequence,
and so do their ancestors (at least in the recent past).

We can calculate pairwise distances between the species once again, but this time using
the branch lengths in the tree.
The distance between two species corresponds to the sum of the branch lengths along the
unique path connecting the two species through their most recent common ancestor.
Because we are dealing with trees (a special type of graph), there is always a unique path
connecting two leaves (so that the notion of "distance" is well defined).

```{r}
parsmny_dist_mat <- cophenetic(tree)
```

You can verify that this matrix is symmetric. Let's visualize this:

```{r}
par(mfrow=c(1,2))
heatmap(parsmny_dist_mat,symm=T)
```

This is slightly different from the pairwise sequence differene matrix.
Let's compare the pairwise distances in both:

```{r}
m1 <- seq_diff_matrix
m2 <- parsmny_dist_mat

m1 <- m1[rownames(m2),colnames(m2)]

m1 <- m1[upper.tri(m1)]
m2 <- m2[upper.tri(m2)]

plot(m1,m2,xlab='Nucleotide differences',ylab='Patristic distances (parsimony)')
```

Clearly these things are correlated - but they are not identical.

Parsimony reconstructions of trees are more biologically grounded than dendograms build from pairwise 
differences, but we can do even better. 

The criterion
to minimize the number of nucleotide changes in parsimony ignores back-substitutions that can undo nucleotide changes.
Even though such events may be rare in the grand scheme of things, over long timescales they can occur. 
Ignoring this can lead to spurious conclusions. 

This is the motivation for the final tree construction method we will consider here -- the so-called
Likelihood methods that explicitly model nucleotide substitution as a stochastic process (specifically,
as a continuous-time Markov chain). These models depend on parameters that must be estimated from data. As 
we will see later, Markov chains that describe changes in nucleotides over time may be adapted to model
changes in other traits over time (such as location).

Because our focus for this workshop is discrete phylogeography, we will not consider methods making use
of continuous data at the tips of trees. It is possible to formulate diffusion models and stochastic
processes such as Brownian Motion or the Ornstein-Uhlenbeck process to model evolution of continuous traits from root to leaves. 
We will not do so here.

Discrete-state Markov models (as well as the Brownian motion models for trait evolution) fall under
the category of Likelihood methods for ancestral character estimation. This is because it is possible to fit
these models to data using standard techniques from statistics that employ likelihood functions.

The code snippet below will estimate parameters of a discrete-state Markov chain
defined on the state space (A,C,G,T) and changing in continuous time.
We will not estimate the tree topology. Estimatng parameters of the Markov chain
will help us determine the branch lengths in the tree that are most consistent with the data (i.e. the sequences).

```{r}
fit <- pml(tree, primates)
fit <- optim.pml(fit, model="F81", control = pml.control(trace=0))


#This will give us also the most likely ancestral states:
anc.ml <- anc_pml(fit)

#Let's look at site 17 in the sequences:
plotAnc(anc.ml, 17)
title("Maximum Likelihood Estimate")
```


Let's compare the parsimony and ML trees:

```{r}
par(mfrow=c(1,2))

plot(tree, main='Parsimony tree')

plot(anc.ml$tree, main='Maximum likelihood tree')
```

The tree topologies (=genealogical relationship) are the same, but the branch lengths are a bit different. Notice, 
in particular, that the ancestors in the Mouse-Bovine-Tarsier-Lemur clade of the maximum likelihood phylogeny are concentrated
together more closely in time than in the parsimony tree.


Let's calculate pairwise distances in this tree:
```{r}
mldists <- cophenetic(anc.ml$tree)

c1 <- mldists[upper.tri(mldists)]
c2 <- parsmny_dist_mat[upper.tri(parsmny_dist_mat)]

plot(c1,c2,xlab='Maximum likelihood distances', ylab='Parsimony tree distances')
```

So the distances between species in the parsimony and Maximum Likelihood tree using the
Markov model for nucleotide changes are pretty tightly correlated, but not identical. Hypothesis: the 
pairwise differences below the diagonal-ish spread of points in the right-hand side of the plot correspond to locations in sequences
that experienced a back-mutation (so they are actually separated from each other by a longer period of time than their
sequences would otherwise suggest).

# Exercises.

## 1. Statistical phylogenetics, in a nutshell

The code example above treats the tree topology (i.e. the geneaology of the organisms) as fixed, and estimates parameters of the Markov chain (which
modifies the branch lengths). In the reconstruction of a Maximum likelihood phylogeny, we would need to evaluate the likelihoods of different
tree topologies as well as branch lengths in order to produce an estimate of the phylogenetic tree. Surprisingly, tree topology can be treated like
a paramter -- all we would need to do is optimize our likelihood function over the space of phylogenetic trees. 
 
Use the nni function on the original parsimony tree to obtain all of the trees that are one
nearest-neighbor interchange away from the parsimony tree. Fit Markov models to each tree and
calculate the likelihood for each tree under the optimized parameters of the Markov model. Then, compare the likelihoods
of the different tree topologies.
Based off of this, do you think it might be possible to estimate the most likely tree, given
the Markov model structure and the sequence data?

## 2.

Load in a different set of sequences from the workshop and try to construct a parsimony tree.
Use a nucleotide substitution model of your choice (HKY, F81, etc.) to fit a Markov model
describing sequence evolution on the parsimony tree. How much do branch lengths change after
fitting the Markov model?

## 3.

Use the read.tree function to load in a tree that you download from someplace. Try plotting it,
and if you can find associated metadata (from Nextstrain, for example) try plotting that on the tree
as well. It can be handy to combine the plotTree and tiplabels functions from phytools for this purpose.



It is optional, but the ggtree package in R also comes in handy for plotting trees and you might consider
installing it.
